---
title: "CaribicusTrees"

editor: visual
knitr: true
bibliography: CaribicusTrees.bib
---

# Caribicus species trees

## Goals

What follows are species trees generated from the filtered SNP dataset from genotyping-by-sequencing (GBS) data.

## Getting started

We are using *R* version (v) 4.5.2 and *RStudio* v 2026.01.0-392 for the analysis. *Quarto* v 1.6.39 for MacOS was used to document the workflow. Some work was performed on the Pitzer cluster at the Ohio Supercomputing Center. Set the working directory in *RStudio* to one where the filtered `vcf.gz` file and the taxon/population designation file are available.

## General methods summary

## Packages

- ape [@ape]

- phytools [@phytools]

  ## IQ-Tree

IQ-Tree version 3

```{r}
#| Label: Load package libraries
library(ape)
library(phytools)
```

Read the treefile from IQ-Tree. The IQ-TREE analysis was run on the filtered SNP dataset converted to phylip format using `vcf2phylip.py`. IQ-TREE requires all sites to be variable when using the ascertainment bias correction (`+ASC`), so it automatically generates a variable sites only phylip file (`Caribicus_iqtree.varsites.phy`) from which the tree is built. IQ-TREE was also run on a FASTA formatted version of the filtered SNP dataset generated using `vcfR2DNAbin` in `vcfR`. The filtered dataset contains 24,398 SNPs, however 9,124 sites contain IUPAC ambiguity codes representing heterozygous genotypes. IQ-TREE treats sites with ambiguity codes as invariant, leaving 15,274 variable sites. As a result both analyses from the phylip and FASTA inputs are effectively built from the same 15,274 variable sites and produce identical topologies and bootstrap support values.

```{r}
#| Label: Outgroup rooted IQTREE with colors
#| fig-width: 8
#| fig-height: 13
# Read the taxa file to assign species names to project numbers
CaribicusTAXA <- read.table("data/taxa/Caribicus_taxa1.txt", sep = "\t", header = TRUE)

ColorPal_CUSTOM <- c("#004949", "#DB6D00", "#D82632", "#290AD8", "#FED439",
                     "#B66DFF", "#006DDB", "#FA9C88", "#924900",
                     "#7BA9CB", "#008600", "#FF6984", "#5C3317", "#1CE6FF")

# Map population codes (alphabetical) to colors — used for tip color lookup
pop_colors <- setNames(ColorPal_CUSTOM, sort(unique(CaribicusTAXA$pop)))

# Separate named vector with formal species names — used for legend only
species_names <- read.table("data/taxa/Caribicus_species_names.txt",
                            sep = "\t", header = TRUE)
pop_colors_legend <- setNames(pop_colors,
                              species_names$formal[match(names(pop_colors), species_names$code)])

Caribicus_IQTREE <- read.tree("data/iqtree/phy/Caribicus_iqtree_varsites.treefile")

# Map each tip label (sample ID) to its population color
tip_cols <- pop_colors[CaribicusTAXA$pop[match(Caribicus_IQTREE$tip.label, CaribicusTAXA$id)]]

plot(Caribicus_IQTREE, cex = 0.9, tip.color = tip_cols)
nodelabels(Caribicus_IQTREE$node.label, cex = 0.6, frame = "none", adj = c(1.2, -0.3))
legend(x = 0, y = 20, legend = names(pop_colors_legend), fill = pop_colors_legend,
       cex = 0.9, bty = "n", xpd = TRUE, text.font = 3)
```

*Last updated: `r format(Sys.time(), '%B %d, %Y %H:%M')`*
