---
title: "CaribicusFst"

editor: visual
knitr: true
bibliography: CaribicusFst.bib
---

# Fixed differences, F~ST~, and Nei's distance

## Goals

What follows is our workflow for estimating pairwise fixed differences, genetic differentiation (F~ST~), and Nei's genetic distance among galliwasp species and populations based on the filtered SNP dataset generated from genotyping-by-sequencing (GBS) data.

## Getting started

We are using *R* version (v) 4.5.2 and *RStudio* v 2026.01.0-392 for the analysis. *Quarto* v 1.6.39 for MacOS was used to document the workflow. Set the working directory in *RStudio* to one where the filtered `vcf.gz` file and the taxon/population designation file are available.

## General methods summary

Pairwise fixed differences were estimated using the `gl.fixed.diff` function in *dartR.base* v 0.98 [@dartR.base] under default parameters (`alpha = 0.05`, `delta = 0.02`) with p-values estimated for each pairwise comparison. Pairwise F~ST~ was estimated using the `stamppFst` function and pairwise Nei's distances were calculated using the `stamppNeisD` function in *StAMPP* v 1.6.3 [@StAMPP]. Both F~ST~ and Nei's distance are displayed as heatmaps with groups ordered from the lowest to highest mean pairwise values using *ggplot2* v 3.5.2 [@ggplot2]. Groups for all analyses are defined by the `pop` column in `Caribicus_pops1.txt`.

## Packages

- vcfR [@vcfR]

- adegenet [@adegenet]

- StaMPP [@StAMPP]

- reshape2 [@reshape2]

- dartR.base [@dartR.base]

```{r}
#| label: Load package libraries
library(vcfR)
library(adegenet)
library(StAMPP)
library(ggplot2)
library(reshape2)
library(dartR.base)
```

## Data files

This analysis requires the following data files.

- `Caribicus.variants.filtered.vcf.gz` The filtered `vcf` file generated in the [SNP filtering and analysis](CaribicusSNPs.html) page.

- `Caribicus_pops1.txt` The text file containing the individual sample identifications and population/taxon designations used for this analysis.

## VCF file

Load the SNP data from the filtered `vcf.gz` file.

```{r}
#| label: Read filtered vcf
CaribicusVCF <- read.vcfR("data/vcf/Caribicus.variants.filtered.vcf.gz")
CaribicusVCF
```

## Taxon and population assignments

Read the population assignments from `Caribicus_pops1.txt`. Subset to only include samples present in the VCF file and summarize the group counts.

```{r}
#| label: Read population assignments
CaribicusPOPS <- read.table("data/taxa/Caribicus_pops1.txt", sep = "\t", header = TRUE)
CaribicusPOPS <- CaribicusPOPS[CaribicusPOPS$id %in% colnames(CaribicusVCF@gt), ]
table(CaribicusPOPS$pop)
```

Verify that the sample IDs in the population file match those in the VCF file.

```{r}
#| label: Verify sample ID match
match_result <- CaribicusPOPS$id %in% colnames(CaribicusVCF@gt)[-1]
match_summary <- table(match_result)
print(match_summary)
not_matching <- CaribicusPOPS$id[!match_result]
print("Samples not matching:")
print(not_matching)
```

## Convert data to genlight format

Convert the VCF object to a genlight object, set ploidy to 2 for these diploid organisms, and assign populations from `Caribicus_pops1.txt`. The `match()` function ensures population assignments are ordered correctly relative to the genlight individual order regardless of sample ordering in the population file.

```{r}
#| label: Convert to genlight
CaribicusGEN <- vcfR2genlight(CaribicusVCF)
ploidy(CaribicusGEN) <- 2
pop(CaribicusGEN) <- CaribicusPOPS$pop[match(indNames(CaribicusGEN), CaribicusPOPS$id)]
```

## Fixed differences

Run a compliance check on the genlight object using the `gl.compliance.check` function in *dartR.base* prior to the fixed differences analysis.

```{r}
#| label: Fixed differences compliance check
CaribicusGENfdiff <- gl.compliance.check(CaribicusGEN)
```

Estimate pairwise fixed SNP differences between all groups using `gl.fixed.diff` under the default parameters (`alpha = 0.05`, `delta = 0.02`) and include a statistical test for each pairwise comparison (`test = TRUE`).

```{r}
#| label: Fixed differences
#| cache: true 
CaribicusFDIFF <- gl.fixed.diff(CaribicusGENfdiff, test = TRUE)
```

The number of fixed differences (`$fd`) and percent fixed differences (`$pcfd`) are combined into a single table displayed as count (percent).

```{r}
#| label: Fixed differences table
fd <- as.matrix(CaribicusFDIFF$fd)
pcfd <- as.matrix(CaribicusFDIFF$pcfd)
combined <- matrix("", nrow = nrow(fd), ncol = ncol(fd), dimnames = dimnames(fd))
for (i in 1:nrow(fd)) {
  for (j in 1:ncol(fd)) {
    if (!is.na(fd[i, j])) {
      combined[i, j] <- paste0(fd[i, j], " (", pcfd[i, j], "%)")
    }
  }
}
knitr::kable(combined,
             caption = "Pairwise fixed differences: count (percent of loci)",
             align = "r")
```

## F~ST~ and Nei's distance

Generate pairwise F~ST~ and Nei's distance estimates from the genlight object. The lower triangle of each resulting matrix is reflected to the upper triangle to produce a symmetric matrix for heatmap visualization.

```{r}
#| label: Estimate Fst and Neis distance
CaribicusFST <- stamppFst(CaribicusGEN, nboots = 100, percent = 95, nclusters = 4)
CaribicusFSTmatrix <- CaribicusFST$Fsts
CaribicusFSTmatrix[upper.tri(CaribicusFSTmatrix)] <- t(CaribicusFSTmatrix)[upper.tri(CaribicusFSTmatrix)]
CaribicusFSTheat <- reshape::melt(CaribicusFSTmatrix)

CaribicusNeiDmatrix <- stamppNeisD(CaribicusGEN, pop = TRUE)
CaribicusNeiDmatrix[upper.tri(CaribicusNeiDmatrix)] <- t(CaribicusNeiDmatrix)[upper.tri(CaribicusNeiDmatrix)]
CaribicusNeiDheat <- reshape::melt(CaribicusNeiDmatrix)
```

Examine the full F~ST~ results including bootstrap p-values.

```{r}
#| label: Print Fst results
CaribicusFST
```

### Pairwise F~ST~ heatmap

Groups are ordered from the lowest to highest mean pairwise F~ST~ value. Values are rounded to three decimal places and displayed in each cell; diagonal cells (self-comparisons) are shown in grey.

```{r}
#| label: Fst heatmap ordered low to high
#| fig-width: 10
#| fig-height: 9
# Order groups from lowest to highest mean Fst
Caribicus_FSTordered <- CaribicusFSTheat
Caribicus_FSTordered$X1 <- reorder(factor(Caribicus_FSTordered$X1),
                                    Caribicus_FSTordered$value, mean, na.rm = TRUE)
Caribicus_FSTordered$X2 <- reorder(factor(Caribicus_FSTordered$X2),
                                    Caribicus_FSTordered$value, mean, na.rm = TRUE)
ggplot(Caribicus_FSTordered, aes(x = X1, y = X2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = ifelse(is.na(value), "", round(value, 3))), size = 3) +
  scale_fill_gradient(low = "white", high = "darkred",
                      name = expression(F[ST]), na.value = "grey90") +
  labs(x = NULL, y = NULL,
       title = expression("Pairwise" ~ F[ST] ~ "(ordered lowest to highest)")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        axis.text.y = element_text(size = 9),
        plot.title = element_text(size = 11))
```

### Pairwise Nei's distance heatmap

Groups are ordered from the lowest to highest mean pairwise Nei's distance. Values are rounded to three decimal places and displayed in each cell; diagonal cells (self-comparisons) are shown in grey.

```{r}
#| label: Neis distance heatmap ordered low to high
#| fig-width: 10
#| fig-height: 9
# Order groups from lowest to highest mean Nei's distance
Caribicus_NeiDordered <- CaribicusNeiDheat
Caribicus_NeiDordered$X1 <- reorder(factor(Caribicus_NeiDordered$X1),
                                     Caribicus_NeiDordered$value, mean, na.rm = TRUE)
Caribicus_NeiDordered$X2 <- reorder(factor(Caribicus_NeiDordered$X2),
                                     Caribicus_NeiDordered$value, mean, na.rm = TRUE)
ggplot(Caribicus_NeiDordered, aes(x = X1, y = X2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = ifelse(is.na(value), "", round(value, 3))), size = 3) +
  scale_fill_gradient(low = "white", high = "darkblue",
                      name = "Nei's D", na.value = "grey90") +
  labs(x = NULL, y = NULL,
       title = "Pairwise Nei's distance (ordered lowest to highest)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        axis.text.y = element_text(size = 9),
        plot.title = element_text(size = 11))
```

*Last updated: r format(Sys.time(), '%B %d, %Y %H:%M')*
